Robot Team III: Scotty Strikes Back

### 2024-05-06 paper/draft-introduction.tex
updating the outline

\documentclass{article}
\usepackage{graphicx} % Required for inserting images
\usepackage[margin=0.75in]{geometry}

@@

Inland water bodies pose a unique challenge for analysis with remote sensing imagery due to their complex spectral characteristics and small-scale spatial variability. Nevertheless, significant effort has driven the development of techniques for estimating optically-active water quality parameters such as chlorophyll-a, blue-green algae, colored dissolved organic matter (CDOM), turbidity, and temperature from these sources\cite{koponen2002lake,ritchie2003remote, bonansea2015using}. Machine learning has proven to be a powerful tool in this domain, with tree-based models, support vector machines, deep neural networks, and others architectures providing data-driven approaches mapping imagery to parameters of interest \cite{thenkabail2018hyperspectral, ghatkar2019classification,sagan2020monitoring}. However, the training and evaluation of these models demands significant volumes of coincident, in situ data. Collecting these datasets is time consuming and expensive, often requiring observations spanning decades \cite{aurin2018remote, ross2019aquasat}.

The characterization of inland water bodies by remote sensing imagery poses a distinct challenge due to their complex spectral characteristics and small-scale spatial variability \cite{ritchie2003remote}. Traditional multi-spectral imagers utilized by these sensing platforms have large pixel sizes, limited spectral bands, and long revisit times.
- Nevertheless, ... 
- Machine learning has proven... 
- However, these techniques demand significant volumes of in situ data...
- To address the

Hyperspectral imaging
- Hyperspectral imaging has emerged as a pivotal technology in this space.
- With their increased spectral resolution, hyperspectral imagers (HSIs) capture the complex chemical, biological, environmental characteristics of reflectance spectra. (add references for specific applications of HSIs). 
- However HSIs alone do not address the limited spatial resolution of

UAV
- Advancements in hyperspectral imaging technology have lead to a continual decrease in their size facilitating their inclusion as the payload of small, highly maneuverable unmanned aerial vehicles (UAVs) such as drones. Consequently, much work has gone into the development of UAV platforms and associated techniques 
- Near-earth HSI addresses spatial, spectral, and temporal limitations of satellite and airborne platforms. UAV-based HSI enable fine-scale mapping \cite{banerjee2020uav}
- Drones can be equipped with HSI and compute to enable rapid generation of spectral indices like the NDVI for applications such as precision agriculture \cite{horstrand2019uav}
- UAV-based HSI can be georectified to centimeter-scales without need for ground control points by using on-board GPS and IMU \cite{arroyo2019implementation}
- UAV-based HSI for turbidity estimation \cite{vogt2016near}
- Polynomial regression models for chlorophyll-a, turbidity \cite{zhang2022selection}
- For this to be usefull, we still need the in situ data. By combining a UAV born HSI with simultanous collection of ground truth data by a reference sensor equipped autonomous boat, we are able to rapidly train ML models to

Unsupervised Methods
- classic unsupervised ML
- endmember extraction and spectral unmixing

This paper...
- our approach is to use the GTM (a fully Bayesian realization of the self organizing map developed by Biship et al.)
- We demonstrate the capabilities using hyperspectral imagery collected at pond in Montague, North Texas. 
- First we utilize the GTM to perform an unsupervised classification of the water-only pixels in order to reveal the centimeter-scale structures present in the pond
- Next we showhow the GTM can be used to identify relevent spectral endmembers by 
    1. using learned GTM class signatures to identity the abundance of algae (relevant for Harmful Algal Bloom detection)
    2. use the GTM class signatures to map the evolution of a rhodamine tracer dye plume

- \cite{koponen2002lake}
- Example: distinguishing species for harmful algal blooms (HABs) and distinguishing types of oil for oil spill monitoring
- Remote sensing used for oil spill analysis: extent and thickness mapping \cite{kokaly2013spectroscopic, leifer2012state}

- Sun glitter remains a key challenge for sensing in the visible portion of the spectrum but multi and hyperspectral imagers have been used for oil spill identification and to identify their impacts on vegetation stress and mortality \cite{fingas2014review, khan2018modern}

Continued development of HSI technology has led to a dramatic reduction in their size enabling their utilize as the payload of small autonomous aerial vehicles (UAVs) (citation). By flying close to the ground UAV-born hyperspectral imagers can produce centimeter-scale imagery while simultaneously limiting the need for complicated atmospheric corrections required for satellite platforms (citation).

- applications include food quality & safety, medical diagnoses, precision agriculture, and forensic document examination \cite{khan2018modern}
- Hyperspectral data were used to distinguish oils by type, e.g. crude, diesel, gasoline, and palm \cite{yang2020characterization}

@@

In our previous work, we introduced a robot team designed to rapidly characterize water composition in novel environments \cite{robot-team-1}. 
Joining UAV-borne HSI together with collection of in-situ reference data by an autonomous boat dramatically reduces the time required to generate .

By carefully coordinating the simultaneous collection of ground-truth water quality data and hyperspectral imagery, this team can rapidly estimate concentrations of optically-active components including colored dissolved organic matter (CDOM), blue-green algae, crude oil, temperature as well

In this paper, we demonstrate how Generative Topographic Mapping (GTM), a Bayesian alternative to the Self Organizing Map, can be utilized to rapidly identify relevant spectral signatures within a scene.

@@

In this study, we demonstrate how Generative Topographic Mapping (GTM), a Bayesian extension the Self Organizing Map, can be utilized to

The method is fully Bayesian, providing an attractive method for the unsupervised classification of hyperspectral imagery.

- First, we apply the GTM to perform an unsupervised classification for all water pixels. 
    - benefit: GTM classes maintain topological relationships in the latent space resulting in a smooth relationship between the obtained classes...
    - ... revealing the small-scale spatial variability within the pond
- Second, we utilize the GTM to identify reflectance spectra corresponding to specific features. 
- Using these extracted features, we are able to map the extent of algae in the pond in addition to the time evolution of a rhodamine dye tracer.

>>>>

\documentclass{article}
\usepackage{graphicx} % Required for inserting images
\usepackage[margin=0.75in]{geometry}
\usepackage{outlines}

@@

Inland water bodies pose a unique challenge to characterization via remote sensing imagery due to their complex spectral characteristics and small-scale spatial variability \cite{koponen2002lake,ritchie2003remote}. Hyperspectral imaging has emerged as a pivotal technology in this space, where the heightened spectral resolution of hyperspectral images (HSIs) enables the inversion of optically-active water quality parameters such as colored dissolved organic matter, chlorophyll-a, blue-green algae, turbidity, and temperature from captured reflectance spectra \cite{brando2003satellite,moses2012estimation}. Machine learning techniques including tree-based models, support vector machines, deep neural networks, and other architectures have proven effective at mapping observed reflectance spectra to parameters of interest \cite{thenkabail2018hyperspectral,keller2018hyperspectral, ghatkar2019classification,sagan2020monitoring}. However, the calibration and evaluation of these models demands a significant volume of coincident, in situ data. Collecting these paired datasets is time consuming and expensive, often relying on rare satellite overpasses spanning decades of collection \cite{aurin2018remote, ross2019aquasat}. This requirement impedes the real-time assessment of water quality change, for example, during oil spills \cite{fingas2014review,fingas2017review}.

The continued development of hyperspectral imaging technology has resulted in significant size reductions enabling the deployment of hyperspectral imagers on unmanned aerial vehicles (UAVs) such as drones. UAVs are not only less expensive than their remote sensing counterparts but also limit the need for complicated atmospheric corrections and enable centimeter scale spatial resolution through low altitude flights \cite{adao2017hyperspectral, arroyo2019implementation, banerjee2020uav}. UAV-based hyperspectral imaging has seen significant development for use in precision agriculture where the ability to rapidly asses spectral indices like the normalized difference vegetation index (NDVI) is a valuable tool for assessing crop health and irrigation efficiency \cite{horstrand2019uav}. Recently, UAV-based hyperspectral imaging has been utilized to asses a variety of optically active water quality parameters such as chlorophyll-a and suspended solids \cite{lu2021retrieval, zhang2022selection}. In our previous work, we introduced a coordinated robot team designed to enable near-real time characterization of water composition by synchronizing UAV-based HSI collection with in situ data collection by an autonomous boat \cite{robot-team-1, robot-team-2}. However, one key

rests with the availability reference sensors for each parameter of interest. That is, the ability to model water quality parameters depends on

In other words, which sensors to deploy depends on prior knowledge of expected contaminants-of-concern.

% 3. Unsupervised methods in remote sensing and HSI

To take complete advantage of acquired HSI, and in order to guide where to deploy reference sensors, the ability to

we need unsupervised methods which can aid in the identification of sources (in order to alert our attention for further investigation) AND to identify relevant spectral signatures corresponding to quantities of interest.

UAV based HSI address the spectral, spatial, and temporal resolution limitations of remote sensing platforms however, the collection of reference data remains the key limitation hindering the real time characterization of inland waters.

In our previous work, we demonstrated how a coordinated approach combining UAV-borne HSI together with autonomous collection in situ data by an Unmanned Surface Vessel (USV) dramatically reduces the time needed to perform such analyses \cite{robot-team-1, robot-team-2}.

Nevertheless, these approaches all presuppose that we already know what to look for- contaminants which we are not equipped to measure in situ can not be estimated in this paradigm.

Therefore, the identification of unique sources purely from their reflectance signatures is of keen interest for two key reasons. First: we want to be able to direct in situ sensor payloads to areas of interest. Second, we desire to identify relevant spectral signatures which can readily be related to parameters of interest and used for source identification.

Endmember extraction and unmixing approaches can leverage the ability of hyperspectral images to distinguish source types. Example: mineral identification from AVRIS data, algae classification for harmful algal bloom detection, and distinguishing types of oils based on slight difference in spectral shapes.

- Describe classical ML methods such as k-means, fuzzy c-means, DBSCAN, etc in remote sensing and HSI. 
    - Key limitation: how do you determine the correct number of classes?
    - This can be very hard to do without a Bayesian interpretation for the model. e.g. you have to rely on heuristics like the Silhouette diagram.
- Describe related problem of endmember extraction and spectral unmixing. Specifically there are geometric and statistical approaches to solving the problem with ML methods a popular choice for the identification of the relevant family of spectral endmembers.
- To date, much of the focus on drone-based HSI platforms has been on supervised regression of parameters OR the determination of spectral indices.
- An ideal unsupervised approach should accomplish three things
    1. Segmentation of captured imagery into a set of distinct classes
    2. Mapping of distinct spatial regions in collected imagery using the identified classes
    3. Simultaneous identification of relevant spectral endmembers

- Hyperspectral data were used to distinguish oils by type, e.g. crude, diesel, gasoline, and palm \cite{yang2020characterization}
- add citation here about harmful algal blooms as examples

@@

% 4. Generative topographic mapping 
Describe the self organizing map (SOM) as an interesting approach due to the ability to 
    - classify data 
    - maintain topographic relationship between classes which can be used to produces a lower dimensional representation of data distribution (i.e. cluster identification)
    - the weight vector for each class can be interpreted as an endmember in the original data space such that common classes can be interpreted as a family of spectral endmembers.
Cite a few uses ofthe SOM for remote sensing and HSI analysis...
However, the original SOM training algorithm relies on heuristics for controlling the updates during training and offers no clear probabilitistic interpreation. The GTM was developed to specifically address this, providing a fully Bayesian alternative to the SOM.

% 5. This study

In this study, we explore the use of the GTM to perform rapid unsupervised classification of hyperspectral imagery captured by

    - The generative topographic mapping is an attractive algorithm 
        - Probabilistic interpretation
        - Topographic relationship maintained between classes (not just orthogonality)
        - clear fitting criterion with Maximum Likelihood Estimation / Maximum A-Posteriori Estimation
    - We use the GTM to analyze hyperspectral imagery acquired by a drone over pond to 
        - Present method for unsupervised characterization of small-scale variability of pond (i.e. identifying regions of interest which can be used to direct robot boat for further instructions) 
        - Use GTM to extract endmember for algae and map the algal distribution near shore (relevant for Harmful Algal Bloom detection)
        - Use GTM to extract endmember for rhodamine dye and use it to mape evolution of dye plume. This is a demonstration of source identification and mapping. Relevant for the identification of pollution events such as oil spills, industrial seepage, etc...

@@

### 2024-05-06 paper/outline.tex
updating the outline

NOTE: We should move the description of the GTM algroithm to the top of the methods section.

\section*{Outline}
\begin{outline}[enumerate]
\1 Introduction
    \2 Characterization of inland water quality via remote sensing is challenging for 4 key reasons:
        \3 Key limitations:
            \4 Spectral resolution (visible and multi-spectral imaging)
            \4 Spatial Resolution (large pixel sizes)
            \4 Temporal Resolution (long satellite revisit times)
            \4 Availability of in situ reference data
        \3 When appropriate data are available, supervised ML approaches can map reflectance from remote sensing imagery to relevant water quality parameters
    \2 Hyperspectral Imaging addresses spectral resolution
    \2 UAV adresses spatial and temporal resolution
    \2 Robot team addresses in situ data... but what do we do if we don't already know which sensors to use? 
    \2 Unsupervised ML, Endmember extraction, and Spectral Unmixing
        \3 k-NN, fuzzy c-means, matrix factorizations, etc.
            \4 Are general but do not provide relationship between classes e.g. is class 1 more similar to class 2 or class 10?
        \3 SOM addresses this by enforcing a topological relationship between clusters. Classes close to each other in the map are more similar than classes far apart
        \3 Similarly, we may also be interested in the spectral characteristics of the learned *classes*, i.e. we want to think of the classes as *endmembers*.
            \4 Traditional approaches rely on assumption of existence of "pure pixels" in the scene
            \4 additionally, we may have to treat number of endmembers as a hyperparameter

    \2. ...in this paper, we explore how the generative topographic mapping can be used to
        \3 enable principled unsupervised classification of hyperspectral imagery by
            \4 Bayesian treatment (as compared to the self organizing map)
            \4 topological ordering with smooth relationship between classes in latent space
        \3 Train GTM on water pixels to obtain unsupervised classification mapping the detailed structure and small-scale variability of pond in Montague, North Texas
        \3 Mapping can be further utilized for smart deployment of robot team (maybe save for discussion section?) i.e. requisition boat to bring reference sensors to "interesting" area
        \3 Using combined dataset including land, water, and tracer dye, we demonstrate endmember extraction
        \3 Using learned spectra together with NS3, we are able to evaluate extent of algae (application to harmful algal bloom mapping)
        \3 Use learned spectra to map extent

\1 Materials and Methods
    \2 Overview paragraph about project, i.e. plan is to use the Generative Topographic Mapping to analyze drone-based hyperspectral imagery collected at a pond in Montague North Texas on 23-November 2021. 
   \2 Autonomous Robot Team
        \3 Describe drone setup
            \4 Alta X Quadcopter
            \4 Resonon Pika SC2 HSI
                \begin{itemize}
                    \item 462 wavelengths per pixel
                    \item 1600 pixels per scan-line
                    \item embedded GPS/IMU
               \end{itemize}
             \4 Intel NUC for on-board compute
   \2 Data Collection
        \3 Describe collection site in Montague, North Texas
        \3 Collection was performed near solar noon to maximize incident light. At this time of year, the maximum solar elevation was $\approx 60^\circ$ so that sun-glint is not as much of an issue. (Cite sun-glint paper here).
    \2 Pre-processing
        \3 Reflectance conversion of radiance imagery using on-board downwelling irradiance spectrometer
        \3 Georectification of collected
        \3 Wavelength selection to $\lambda \leq 900 \text{nm}$ as longer wavelengths are noisy
        \3 Rescaling of spectra so that peak has $\tilde{R}  = 1.0$  to account for variability in incident lighting conditions due 
    \2 Generative Topographic Mapping
        \3 Start off with Kohonen SOM paper. Key benefit:
            \4 Topological relationship between classes i.e. similarity vs dissimilarity as a function of distance in mapping space (latent space)
        \3 Limitations of SOM
            \4 Not probabilistic
            \4 No clear convergence criterion
            \4 Heuristic based approach with no clear way to select hyperparameters like neighborhood function, number of classes, etc...
            \4 Heskes did reinterpret algorithm as stochastic gradient descent on a cost function...
        \3 GTM Algorithm
            \4 Describe Algorithm
            \4 Describe parameters
        \3 Bayesian information criterion and Akaike information criterion for hyperaprameter optimization
            \4 As an alternative to performing full Bayesian treatment for marginalization over hyperparameters
        \3 Abundance mapping with spectral similarity score
            \4 paper comparing different distance functions for spectra
                \begin{itemize}
                    \item curse of dimensionality can be pernicious here as we have > 462 wavelengths per pixel
                \end{itemize}
            \4 NS3 offers good trade off between spectral angle and euclidean-like distance (MSE)
        \3 Describe use of thresholded NS3 for mapping abundance of GTM derived endmembers.
    \2 Overview of two approaches
        \3 GTM trained on water-only pixels for unsupervised classification of pond
            \4 Goal: Reveal the small-scale spatial variability of pond water
            \4 Discuss how unsupervised classification can help inform autnomous data collection by robot team (save for discussion?)
        \3 GTM trained on combined dataset including water, land, and dye plume pixels
            \4 Fit GTM
            \4 Identify spectral signatures of key exemplar spectra (endmembers)
            \4 Use endmembers to map algal abundance and time evolution of rhodamine dye plume
\1 Results
    \2  GTM fit on Full Dataset
    \2 Hyperparameter optimization
    \2 Spectral signature identification
    \2 Water Class Map (water-only GTM)
    \2 Algae Identification
    \2 Dye plume identification

\1 Discussion
    \2 Overview of GTM applications in our paper
        \3 Unsupervised classifciation of entire pond (no-dye)
        \3 Endmember extraction of spectral signatures
    \2 Comparison to existing approaches
        \3 unsupervised classification methods (to general, no relationship between classes) 
        \3 endmember extraction (mostly linear, assume existence of pure spectra)
        \3 deep learning approaches to endmember extraction don't provide clear way to asses number of endmembers needed
    \2 Applications of either of these with drones?
        \3 Are there any papers using unsupervised methods or endmember extraction with drone based imagery (HSI or otherwise)?
    \2. Use of GTM as component of Robot Team
        \3 As a feature transformer / preprocessing step for supervised methods (as opposed to PCA, k-NN, etc...)
        \3 identification of "interesting" regions for smart provisioning of robot boat
            \4 Standard approach is to survey in a grid
            \4 Can use classification map with real-time measurement to ensure optimal collection of data by collecting samples with from as many GTM classes as possible
    \2 Extensions and future work
        \3 Incorporation of GTM extensions for efficient online-learning
            \4 Comment on time to train standard GTM
                \begin{itemize}
                    \item Key parameters are size of dataset (number of records) and number of RBF centers, $M$.
                \end{itemize}
            \4 Bishop suggest extension for on-line training of GTM using batches.
            \4 All of this could be done in the field limiting the time and cost of surveying as well as optimizing the time-to-insights
        \3 Utilize GTM with further in situ data collection to identify spectral signatures of specific algal species for HAB detection
        \3 Same but for crude oil types...

\1 Conclusions

@@

>>>>

\section*{Introduction}
\begin{outline}[enumerate]
\1 Challenges for water quality characterization
    \2 Traditional remote sensing suffers from limited spatial, spectral, and temporal resolution
        \3 Papers: 
            \4 Broad bands of traditional remote sensing multi-spectral imagers struggle to discern water quality parameters like chlorophyll-a in highly turbid waters due to domintation of spectral signal by sediments. \cite{ritchie2003remote}
            \4 Small, irregular shape of Finnish lakes makes analysis using medium resolution satellite data challenging due to reflectance from shore and near-shore vegetation. "If even a small protion of a pixel is covered by land the retrieval of water quality variables may not be possible" \cite{koponen2002lake}
    \2 This can be addressed by combining two technologies: 
        \3 Hyperspectral Imaging (improved spectral resolution)
        \3 Unmanned Aerial Vehicles (improves spatial and temporal resolution)
        \3 Studies using UAV + HSI for water quality: "UAV-based HSI has emerged as a promising tool..."
            \4 UAV-based HSI addresses resolution, signal-to-noise, and cloud cover issues that plague traditional satellite and high-altitude platforms\cite{banerjee2020uav}
            \4 UAV-based HSI for ecological monitoring \cite{arroyo2019implementation}
    \2 Remaining challenge: how can we extract meaningful insights from complicated spectral data?
        \3 "Ability for measuring hundreds of bands raises complexity when considering the sheer quantity of acquired data, whose usefulness depends on both calibration and corrective tasks occruing in pre- and post-flight stages" \cite{adao2017hyperspectral}
\1 Current approaches for characterization
    \2 Band ratios and ML for water quality parameter inversion
        \3 Papers:
            \4 UAV-based HSI with embedding processing capabiltieis enables rapid generation of spectral indices like the NDVI \cite{horstrand2019uav}
            \4 UAV-based multispectral imaging used to map turbidity (Secchi Disk Depth) \cite{vogt2016near}
            \4 Polynomial combinations of HSI bands from UAV-mounted HSI imagery used to map turbidity, total nitrogen, total phosphorus, ammonia nitrogen, permanganate index, and chemical oxygen demand \cite{zhang2022selection}
            \4 ML methods used to map UAV-based HSI to chlorophyll-a and suspended solids. Tree-based methods worked best (Catboost) \cite{lu2021retrieval}
            \4 Estimation of CDOM, Chlorophyll-a, turbidity, diatoms, and green-algae from HSI using Machine Learning with hyperspectral sensor and two in-situ sensor probes \cite{keller2018hyperspectral}
            \4 ICA and SVM used for HSI classification (not water) \cite{liang2010hyperspectral}
            \4 SVA and LDA used for HSI classification (not water) \cite{zhangSVM2014}
    \2 Key limitation: need for in situ reference data
        \3 For remote sensing applications, collection of datasets relies on serendipitous satellite overpasses. Acquiring sufficient volumes of data needed to calibrate and evaluate inversion models can require decades of observations \cite{aurin2018remote,ross2019aquasat}
        \3 Can be addressed by extending the UAV approach to coordinated robot teams including autonomous in situ sensing \cite{robot-team-1, robot-team-2}
    \2 Remaining limitation: insights are limited by available reference sensors 
        \3 Sensor require careful calibration, for example using specific standards which may not be relevant for local water
    \2 Need: unsupervised methods to extract meaningful information from full HSI data
        \3 Useful for deciding where/how to collect reference data
        \3 Useful for time-critical applications where identification of anomolous sources is important e.g. oil spills, industrial waste, etc... 
\1 Unsupervised ML Approaches
    \2 Dimensionality reduction
        \3 PCA is a popular choice but is a linear model
            \4 Mineral endmembers identified from HSI using sparse PCA \cite{yousefi2016mineral}
        \3 t-SNE can be computationally complex 
    \2 Unsupervised Classification
        \3 k-NN, fuzzy c-means, DBSCAN, etc... matrix factorizations
        \3 Can be helpful for identifying clusters, but don't aid in visualization by themselves
            \4 Robust Manifold Matrix Factorization used for dimensionality reduction \textit{and} clustering \cite{zhang2019hyperspectral}
    \2 Endmember extraction and Spectral Unmixing
        \3 Geometric and statistical methods
        \3 Can be useful for identifying relevant endmembers but may rely on presence of pure pixels
            \4 This assumption may not be valid for water leafing reflectance spectra due to interactions of suspended solids, dissolved components, etc...
        \3 Once endmembers are identified, abundance maps can be produced using spectral angle mapper (e.g. cosine distance) or unmixing models
            \4 However, this does not itself to visualization
    \2 Need: a (principled) method which simultaneously enables dimensionality reduction \textit{and} classification/endmember extraction.
\1 Self Organizing Map and Generative Topographic Mapping
    \2 Self Organizing map developed by Kohonen based on principle that \textit{neurons that fire together, wire together}.
        \3 Construct an unsupervised classification into $N=n\times n$ classes
        \3 Classes are arranged in a regular grid of typically two dimensions with different possible topologies:
            \4 square vs hexagonal lattice (number of neighbors)
            \4 rectangular, cylindrical, toroidal topologies based on boundary conditions
        \3 The map is trained to enforce similarity between neighboring nodes, that is, classes near each other in the SOM grid are more similar than nodes that are far apart. 
        \3 In the context of HSI and water quality, the SOM grid can be used to perform dimensionality reduction via the location of each datapoint in the SOM grid 
        \3 SOM nodes can \textit{also} be interpreted as endmembers/classes
            \4 Add citations here, e.g. David's dust sources paper and other SOM related HSI studies
            \4 SOM used for endmember extraction with HSI for water quality \cite{cantero2004analysis}
    \2 Drawbacks of SOM
        \3 Training algorithm is based on heuristic approach
        \3 Does not provide a clear probabilistic interpretation or a principled approach for hyperparameter choice
    \2 The Generative Topographic Mapping addresses these issues
        \3 Fully Bayesian re-interpretation of the SOM
        \3 Enables both dimensionality reduction \textit{and} classification like the SOM
        \3 Uses of the GTM in HSI:
\1 Paper overview
    \2 Development of the GTM as an efficient approach for dimensionality reduction and unsupervised classification of UAV-acquired HSI
    \2 Apply the GTM to data collected at a pond in Montague, North Texas
    \2 Train a GTM on water-only pixels to provide detailed segmentation water spectra
    \2 Use GTM on combined dataset for demonstration of endmember extraction and source mapping
        \3 Algal abundance (i.e. application to algal bloom)
            \4 Comment on potential application for algal species discrimination
        \3 Rhodamine Dye Tracer simulating dispersion of pollution source
            \4 Application to oil spill identification

@@

\section*{Materials and Methods}
\section*{Results}
\section*{Discussion}
\section*{Conclusions}

### 2024-05-06 paper/paper.tex
updating the outline

1. Introductory paragraph

Inland water bodies pose a unique challenge to characterization via remote sensing imagery due to their complex spectral characteristics and small-scale spatial variability (cite reviews here). Hyperspectral imaging has emerged as a pivotal technology in this space, where increased spectral resolution enables the inversion of optically-active water quality parameters such as colored dissolved organic matter (CDOM), chlorophyll-a, blue-green algae, turbidity, and temperature from captured reflectance spectra \cite{koponen2002lake,ritchie2003remote, bonansea2015using}. Machine learning has proven to be a powerful tool in this domain, with tree-based models, support vector machines, deep neural networks, and others architectures providing data-driven approaches to map observed reflectance spectra to parameters of interest \cite{thenkabail2018hyperspectral, ghatkar2019classification,sagan2020monitoring}. However, the training and evaluation of these models demands significant volumes of coincident, in situ data. Collecting these datasets is time consuming and expensive, often requiring observations spanning decades \cite{aurin2018remote, ross2019aquasat}.

2. HSI equipped drones + in situ reference sensing --> still require that we know what to look for.

due to the dramatic improvement in spectral resolution afforded by HSI platforms. Specifically, the combination with hyperspectral imaging together with autonomous collection by UAVs enables unparalleled spectral, spatial, and temporal resolution. Significant effort has developed the use of hyperspectral imagery for the inversion of optically-active water quality parameters. However, this approach requires access to sufficient quantities of in situ reference data to fit and evaluate each model. Traditionally, this process has required collocation of remote sensing imagery with data from field campaigns spanning decades (cite Aurin and Ross). Recently, we have demonstrated how combining autonomous collection of in situ data with UAV-borne hyperspectral imagery dramatically reduces the time required to collect these data. Both of these approaches require we already know what to look for.

Nevertheless, significant effort has driven the development of techniques for estimating optically-active water quality parameters such as chlorophyll-a, blue-green algae, colored dissolved organic matter (CDOM), turbidity, and temperature from these sources.

To take complete advantage we need unsupervised methods which can aid in the identification of sources (in order to alert our attention for further investigation) AND to identify relvant spectral signatures corresponding to quantities of interest. UAV based HSI address the spectral, spatial, and temporal resolution limitations of remote sensing platforms however, the collection of reference data remains the key limitation hindering the real time characterization of inland waters. In our previous work, we demonstrated how a coordinated approach combining UAV-borne HSI together with autonomous collection in situ data by an Unmanned Surface Vessel (USV) dramatically reduces the time needed to perform such analyses \cite{robot-team-1, robot-team-2}. Nevertheless, these approaches all presuppose that we already know what to look for- contaminants which we are not equipped to measure in situ can not be estimated in this paradigm. Therefore, the identification of unique sources purely from their reflectance signatures is of keen interest for two key reasons. First: we want to be able to direct in situ sensor payloads to areas of interest. Second, we desire to identify relevant spectral signatures which can readily be related to parameters of interest and used for source identification.

3. Unsupervised methods in remote sensing and HSI
- Describe classical ML methods such as k-means, fuzzy c-means, DBSCAN, etc in remote sensing and HSI. 
    - Key limitation: how do you determine the correct number of classes?
    - This can be very hard to do without a Bayesian interpretation for the model. e.g. you have to rely on heuristics like the Silhouette diagram.
- Describe related problem of endmember extraction and spectral unmixing. Specifically there are geometric and statistical approaches to solving the problem with ML methods a popular choice for the identification of the relvant family of spectral endmembers.
- To date, much of the focus on drone-based HSI platforms has been on supervised regression of parameters OR the determination of spectral indices.
- An ideal unsupervised approach should accomplish three things
    1. Segmentation of captured imagery into a set of distinct classes
    2. Mapping of distinct spatial regions in collected imagery using the identified classes
    3. Simultaneous identification of relevant spectral endmembers

4. Generative topographic mapping 
Describe the self organizing map (SOM) as an interesting approach due to the ability to 
    - classify data 
    - maintain topographic relationship between classes which can be used to produces a lower dimensional representation of data distribution (i.e. cluster identification)
    - the weight vector for each class can be interpreted as an endmember in the original data space such that common classes can be interpreted as a family of spectral endmembers.
Cite a few uses ofthe SOM for remote sensing and HSI analysis...
However, the original SOM training algorithm relies on heuristics for controlling the updates during training and offers no clear probabilitistic interpreation. The GTM was developed to specifically address this, providing a fully Bayesian alternative to the SOM.

5. This study

In this study, we explore the use of the GTM to perform rapid unsupervised classification of hyperspectral imagery captured by

@@

\subsection{Generative Topographic Mapping}

@@

@@

>>>>

@@

In this study, we explore the use of Generative Topographic Mapping as a tool for the unsupervised classification and nonlinear endmember extraction of hyperspectral imagery. To this end, HSIs were collected using a USV mounted hyperspectral imager at a pond in Montague, North Texas on on 23 November 2020 (shown in panel (d) of Fig~\ref{fig:sample-spectra}). In this section we shall first provide a detailed description of the GTM algorithm with subsequent subsections reserved to describe the specific data collection, processing, and analysis techniques employed in the study.

%%% --- Generative 0Topographic Mapping ------------------------------------
\subsection{Generative Topographic Mapping}

The Generative Topographic Mapping (GTM) was developed by Bishop, Svens\'en, and Williams as a Bayesian realization of Kohonen's Self-organizing Map (SOM) \cite{gtm-bishop-1}. Specifically, the

@@

As outlined in the introduction, an effective unsupervised classification algorithm for hyperspectral imagery should accomplish the following three things: 
\begin{enumerate}
    \item Segmentation of captured imagery into a set of distinct classes
    \item Mapping of distinct spatial regions in collected imagery using the identified classes
    \item Simultaneous identification of relevant spectral endmembers
\end{enumerate}

@@

- Drones can be equipped with HSI and compute to enable rapid generation of spectral indices like the NDVI for applications such as precision agriculture \cite{horstrand2019uav}
